# MACS 40700 Data Visualization - Final Project

For my final project in this course, I created a portfolio to visualize my MA thesis results. Below, I present the plots I developed using R Studio, ggplot2, Shiny, and Tableau. Taking this course allowed me to:

1)  Gain hands-on experience with data visualization tools.
2)  Understand data visualization principles such as truthfulness, functionality, beauty, insightfulness and enlightenment.

Accompanying the plots, you can also look at the rationale of my design choices and intended takeaways.

<br><br>

::: {style="text-align: center;"}
## Overview of MA thesis project

### The Moralization of Self-Improvement
:::

Self-improvement is meant to help people grow, being an “essential human calling” —according to philosopher Rudolf Allers— to achieve our fullest potential (García-Alandete 2022). However, in today’s culture the desire to improve has often been portrayed with a moralized tone that has been related to increased anxiety, pressure, and self-doubt. While self-improvement has been studied by philosophers and psychologists since long ago, the way it is experienced today requires a specific examination in its current context (Coeckelbergh 2022).

The intrinsically positive pursuit of self-improvement and choosing goals that align with one’s ideals has turned into a source of pressure and stress when these goals are framed as socially expected moral obligations. Discussions around self improvement today do not simply encourage growth but frame certain specific behaviors, that are sometimes unattainable for many, as moral imperatives. This can lead to the stigmatization of those who do not conform to what is perceived as “correct” or “morally right”. Furthermore, the new standards transform self-care into a source of guilt (Kraaijeveld and Jamrozik 2022).

Previous research has analyzed the use of moralizing language on social media platforms, as well as users’ experiences when engaging with content. For instance, one study identified moralization in Instagram influencers’ discourse around self-improvement, where they promote very specific behaviors as the correct (Baker 2022). Other studies have shown that engagement with Instagram trends focused on self-improvement (such as #fitspiration for physical appearance) can lead to self-obsession, perfectionism, stress, feelings of inadequacy, and even depression (Coeckelbergh 2022; Bell, Talbot, and Deighton-Smith 2024). While these studies have either focused on the moralizing tone or the emotional response, less is known about the relationship between the two specifically: how moral language itself contributes to the emotional experience of self-improvement. Additionally, most of these studies have been small-scale and have largely concentrated on physical appearance goals, leaving a gap in understanding how these patterns manifest on a broader scale or in relation to other areas of self-improvement.

This research aimed to address that gap by analyzing the presence of moral language in self-improvement discussions at a large scale and across different sub-topics, as well as evaluating its relationship with negative emotions. It was found that moralization language does exist around the desire to self improve and that content with higher moralization language tends to show more negative emotion. By examining self-improvement through these lens, this study contributes to a deeper understanding of how human aspirations for growth are shaped by the cultural narratives around them that frame them as moral obligations. Ultimately, this project is only one step of a larger research agenda: How can the human tendency for self-improvement be experienced in a way that leads to a truly "good life" rather than unattainable ideals?

::: {style="text-align: center;"}
<br><br>

### Data
:::

Data from the subreddit [r/selfimprovement](https://www.reddit.com/r/selfimprovement/) was obtained through [Academic Torrents.](https://academictorrents.com/details/56aa49f9653ba545f48df2e33679f014d2829c10)

::: {style="text-align: center;"}
<br><br>

### Guiding Research Questions
:::

Three research questions were formulated to guide the project:

-   RQ 1. Is moral language present when talking about self-improvement?
-   RQ 2. Are there some specific topics within self-improvement discussions that tend to be discussed with a higher moral tone?
-   RQ 3. How does the usage of moral language associate with emotions in the context of self-improvement?

The following visualizations are organized by research question.

<br><br>

::: {style="text-align: center;"}
### 1. Moralization across subreddits

<iframe src="https://public.tableau.com/views/RQ1_17419062265120/Dashboard1?:showVizHome=no&:embed=true" width="600px" height="800px"></iframe>
:::
<br>

For assessing the presence of moral language in r/selfimprovement, it was compared with other two subreddits: r/investing and r/homeowners. Three different ways of measuring moralization were employed:

1- The moral dimension from the [LIWC dictionary](https://www.liwc.app/)

2- The [Moral Foundations Dictionary 2.0 (MFD)](https://www.liwc.app/dictionaries) (also using the LIWC software)

3- [MoralBERT](https://github.com/vjosapreniqi/MoralBERT)

The goal of the visualization was to present the results from the three methods collectively, thus strengthening the argument that self-improvement is a more moralized topic compared to others, regardless of the method for measuring employed.

Using bar charts is a very simple approach but it makes the visualization easy to interpret and therefore functional. The scores, according to each method, are displayed on the axes to allow the user to visualize the raw score from the measurement tool, staying truthful to the data.

The Y-axes for the top plot are independent because the MFD scores tend to be higher, as it incorporates multiple dimensions, while the MoralBERT is also independent since its results are calculated in a fundamentally different way from the dictionary-based approaches. An alternative presentation could be through tables, but this would make it difficult to grasp the overall picture, given that each measurement operates on a different scale. The visualization, therefore, adds insight by combining the findings into a single narrative: that self-improvement is a more moralized topic. Ultimately, it is enlightening, as it visually highlights the gap in moralization compared to other topics, encouraging the reader to reflect on the gap and what it means.

Both the MFD and the MoralBERT results are based on the [Moral Foundations Theory](https://moralfoundations.org/) that suggest that morality is composed of different “foundations” or values, so I chose stacked bar charts to separate between each foundation. By excluding foundation labels, the emphasis remains on the overall score and simplicity is preserved. However, interactivity allows users to hover over each foundation to explore the specific foundations details. This interactivity is precisely why Tableau was chosen for the visualization. The color choices for the MFD and MoralBERT results were carefully selected to represent each foundation's values while maintaining accessibility. Both used consistent colors to make it visually appealing. The chosen colors aim to evoke associations with the underlying concepts of each foundation:

-   Authority (Red): Red is often associated with power, strength, and leadership.
-   Care (Light Blue): Light blue is commonly linked to calmness, trust, and compassion, aligning with the Care foundation’s focus on empathy, nurturing, and concern for others’ well-being.
-   Fairness (Yellow): Yellow represents fairness as a value tied to justice and equality
-   Loyalty (Orange): Orange conveys warmth, enthusiasm, and connection, reflecting the sense of loyalty, unity, and community that this foundation promotes.
-   Sanctity (Light Purple): Light purple evokes a sense of purity and spirituality.

The LIWC bars are more straightforward, representing a single score and thus using only one color. Hovering over them reveals the mean score values for each subreddit. The combination of the military green with the moral foundation’s colors create harmony for maintaining beauty.

<br><br>

::: {style="text-align: center;"}
### 2. Moralization within self-improvement topics
:::

```{r topics bubbles plot, message = F, echo = F, warning = F}
library(ggplot2)
library(dplyr)
library(forcats)
library(ggtext)
library(showtext)

data <- read.csv("datavis-files/data/final_data.csv")
```

```{r topic bubbles data preparation, message = F, echo = F, warning = F}

average_morality <- data %>% 
  group_by(dominant_topic) %>%
  summarise(
    average_moral_liwc = mean(moral),
    average_mfd = mean(Foundations_total_score),
    average_Care_Virtue = mean(Care_Virtue),
    average_Care_Vice = mean(Care_Vice),
    average_Fairness_Virtue = mean(Fairness_Virtue),
    average_Fairness_Vice = mean(Fairness_Vice),
    average_Loyalty_Virtue = mean(Loyalty_Virtue),
    average_Loyalty_Vice = mean(Loyalty_Vice),
    average_Authority_Virtue = mean(Authority_Virtue),
    average_Authority_Vice = mean(Authority_Vice),
    average_Sanctity_Virtue = mean(Sanctity_Virtue),
    average_Sanctity_Vice = mean(Sanctity_Vice),
    average_Care_total = mean(Care_total),
    average_Fairness_total = mean(Fairness_total),
    average_Loyalty_total = mean(Loyalty_total),
    average_Authority_total = mean(Authority_total),
    average_Sanctity_total = mean(Sanctity_total),
    count = n()
  )

# Create a named vector to map topic numbers to real labels
topic_mapping <- c(
  "Topic 1" = "Habits & Goals",
  "Topic 2" = "Education & Career",
  "Topic 3" = "Social life",
  "Topic 4" = "Health & Fitness",
  "Topic 5" = "Social media & Entertainment",
  "Topic 6" = "Confidence & Dating",
  "Topic 7" = "Family & Finances",
  "Topic 8" = "Learning",
  "Topic 9" = "Mental health"
)

# Replace topic names
average_morality <- average_morality %>%
  mutate(topic_name = recode(dominant_topic, !!!topic_mapping)) %>%
  # reorder so topics with more morality appear at the top
  mutate(topic_name = fct_reorder(topic_name, average_moral_liwc, .desc = FALSE))

color_topics = c(
        "Habits & Goals" = "#E69F00",
        "Education & Career" = "#0072B2",
        "Social life" = "#56B4E9",
        "Health & Fitness" = "#009E73",
        "Social media & Entertainment" = "#F0E442",
        "Confidence & Dating" = "#D55E00",
        "Family & Finances" = "#CC79A7",
        "Learning" = "#999999",
        "Mental health" = "#660099")

reference_mean = 0.25
```

```{r topic bubbles plot, message = F, echo = F, warning = F}

fig1 <- ggplot(average_morality, aes(x = average_moral_liwc, y = factor(topic_name), size = count, color = topic_name)) +
  geom_point(alpha = 0.6, position = position_jitter(height = 0.2, width = 0)) +  
  geom_vline(xintercept=reference_mean, color="red", linetype ="dashed",size=0.5) +
  scale_size_continuous(range = c(10, 25)) +
  scale_color_manual(values = color_topics) +  
  labs(title = "Moral Expression Across Topics",
       subtitle = "(Size proportional to percentage of corpus)",
       x = "Avg. Moral LIWC Score compared to <span style='color:red;'>- -LIWC testing corpus baseline- -</span>",
       y = "Topics") +
  theme_bw() +
    theme(
    plot.title = element_text(size = 20, hjust=0.5),  
    plot.subtitle = element_text(size = 10, hjust=0.5),
    axis.title.x = element_markdown(size = 10),            
    axis.title.y = element_text(size = 10),
    axis.text.x = element_text(size = 9),              
    axis.text.y = element_text(size = 9),
    legend.position = "none")

fig1
```

<br><br>
A bubble chart was used to incorporate three dimensions: the topics within the r/selfimprovement, the average moral LIWC score for documents in each topic, and the proportion of each topic within the corpus. While the proportion isn’t central to the research question, varying bubble sizes make the visualization more engaging for beauty.


Placing the moral scores on the horizontal axis frames morality as a spectrum, allowing readers to easily compare topics in terms of their average moral tone. Additionally, a reference line was included based on the LIWC manual’s reported mean for morality scores across Reddit and other sources. This helps contextualize the results and make it functional, showing how some topics are particularly more moralized than others. 

Using raw LIWC scores on the x-axis also keeps the plot transparent and grounded in the data to preserve truthfulness.

Similar to the RQ1 plots, the colors where carefully chosen to be color-blind friendly and create harmony. 

One key takeaway from the visualization is that only the “social topics” exceed the baseline moral score, an insight facilitated by the plot. This pattern would be less apparent without sorting and comparing topics against a reference point. The plot is enlightening as it invites interpretation, hinting that something unique may be happening with these socially-oriented discussions, and it encourages the reader to explore why that might be.

<br><br>

<iframe src="https://natashacarpcast.shinyapps.io/shiny-app/" style="border: 1px solid #AAA; width: 800px; height: 900px">

</iframe>

After visualizing the difference in general moralization among the topics, this other interactive plot uses a Shiny app to allow the user to explore the different moral foundations proposed by the [Moral Foundations Theory](https://moralfoundations.org/) and see how the topics vary in scores among those. 


This visualization not only shows the overall moral foundation score for each topic, but it also includes two additional subplots that break down each foundation into its virtue and vice components. According to the [Moral Foundations Theory](https://moralfoundations.org/), moral language can move in two directions: a virtue tone, which praises positive behaviors or attitudes (e.g., honest), and a vice tone, which conveys blame or negativity (e.g., dishonest). Including both the combined and separated scores allows the reader to gain deeper insights into how moral language is used within each topic. For instance, the “Health and Fitness” topic stands out for having the highest level of sanctity-related language, and notably, this is more strongly expressed through virtue terms than vice.


The use of an interactive plot helps to present a lot of information without it being visually overwhelming or repetitive (as it would be if they were all displayed on the webpage at the same time).  Presenting each foundation separately but within the same interactive space lets the reader explore the data in an engaging way. 


A key design choice was to fix the y-axis across all plots: for total, virtue, and vice scores across every foundation. By default, ggplot adjusts the y-axis based on the data range, so differences between foundations or vice/virtue plots would not be easy to grasp in such way. Fixing the axis prevents this distortion and makes comparisons across foundations more accurate and intuitive. For example, the “fairness” tab displays shorter bars, but this visually shows the insight that this foundation simply plays a smaller role in the dataset compared to other foundations like “care.”


While the research question focuses on topics and not foundations, comparing between foundations adds an important nuance while still comparing among topics too. It highlights how different values are distributed across self-improvement discourse and reveals which moral themes are most prominent within which specific topics. After exploring the five moral dimensions, it is enlighting to visualize the patterns of the interaction between topics and foundations, leaving a lot to reflect on our current understanding of self improvement. 

<br><br>

::: {style="text-align: center;"}
### 3. Moralization and emotions

<iframe src="https://public.tableau.com/views/RQ3-1/Dashboard3?:showVizHome=no&:embed=true" width="600px" height="750px"></iframe>

:::


The goal of this visualization is to highlight differences in negative emotion, both general and anger-specific, between posts that are moralized and those that are not. A bar plot was chosen for its simplicity and clarity, making it easy and functional for readers to compare values at a first look. The y-axis is clearly labeled, with an explicit title that accurately reflects the meaning of the values, preserving transparency and truthfulness in the data presentation.

Colors were selected carefully: blue represents general negativity and red represents anger. This combination is not only intuitive but also visually harmonious, enhancing the visual appeal of the plot.

By dividing the corpus into two distinct groups—moral vs. non-moral, based on the LIWC’s reported mean, the plot offers a straightforward way to explore the relationship between moral language and negative emotion. This binary approach simplifies the comparison and avoids complications from data imbalance, which might otherwise limit the usefulness of boxplots or scatterplots. In doing so, it reinforces the functionality of the visualization.

Overall, the plot communicates that posts with higher moral tone tend to contain more negative emotion, supporting the hypothesis that moralization may be linked to emotional tension. It offers a compelling insight that contributes to broader discussions around the emotional dynamics of moral language.

::: {style="text-align: center;"}
<iframe src="https://public.tableau.com/views/RQ3-2/Dashboard2?:showVizHome=no&:embed=true" width="620" height="600"></iframe>

:::

This final plot explores the nuances of moral language at the word level. Built with Tableau, it allows for interactivity: users can hover over individual data points to see the specific word and the topic it belongs to, making the experience more engaging and informative.
Words are sized according to their frequency and colored by topic. If a word appears in multiple topics, it is repeated accordingly, which enables a more detailed exploration of how moral language shifts across contexts. For instance, the word “wrong” takes on a noticeably stronger negative tone in the Mental Health topic than in Confidence & Dating, highlighting how emotional connotation can vary by theme.
Using a text data visualization revealed two interesting insights that would have been impossible to detect without a plot:
As words become more negative in emotional tone, they also tend to become more positive, suggesting that self-improvement may often carry a dual emotional charge, capturing both the uplifting and challenging aspects of personal growth.
Words cluster spatially by topic across the emotional axes, indicating that topic influences the emotional tone of moral language.

Although the plot includes a large number of words, making it difficult to interpret every one individually, these broader patterns offer meaningful insights. They add depth to our understanding of how morality is expressed across different areas of self-improvement, inviting further reflection on the emotional complexity of moralized discourse.


## References

Baker, Stephanie Alice. 2022. “Alt. Health Influencers: How Wellness Culture and Web Culture Have Been Weaponised to Promote Conspiracy Theories and Far-Right Extremism During the COVID-19 Pandemic.” EUROPEAN JOURNAL OF CULTURAL STUDIES 25 (1): 3–24. <https://doi.org/10.1177/13675494211062623>.

Bell, Beth T., Catherine V. Talbot, and Nova Deighton-Smith. 2024. “Following up on #Fitspiration: A Comparative Content Analysis and Thematic Analysis of Social Media Content Aiming to Inspire Fitness from 2014 and 2021.” PSYCHOLOGY OF POPULAR MEDIA, February. <https://doi.org/10.1037/ppm0000523>.

Coeckelbergh, Mark. 2022. Self-Improvement : Technologies of the Soul in the Age of Artificial Intelligence. No Limits. New York: Columbia University Press. <https://research.ebsco.com/linkprocessor/plink?id=d6c03c14-eee3-3085-a20e-3cdf5008a557>.

García-Alandete, Joaquín. 2022. “Striving for Personal Perfection: Rudolf Allers’s Psycho‑ethic‑metaphysical Approach to Character and Self‑improvement.” Journal of Religion and Health 61 (4): 3363–83. <https://doi.org/10.1007/s10943-022-01605-6>.

Kraaijeveld, Steven R., and Euzebiusz Jamrozik. 2022. “Moralization and Mismoralization in Public Health.” Medicine, Health Care and Philosophy 25 (4): 655–69. <https://doi.org/10.1007/s11019-022-10103-1>.
